Arrows for Parallel Computations

### 2017-07-03 paper/content/motivation.tex
added olegl changes

%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is just yet another arrow combinator. The resulting arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual |ArrowParallel| instance.
\end{itemize}

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel |map| with and without load balancing, |map-reduce|) in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (|pipe|, |ring|, |torus|). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

>>>>

%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator. The resulting Arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual |ArrowParallel| instance.
\end{itemize}

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours (Sec.~\ref{sec:parEvalNIntro}) and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons  in our newly defined dialect: parallel |map| with and without load balancing. More advanced skeletons are showcased in Section~\ref{sec:topology-skeletons} (|pipe|, |ring|, |torus|). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

### 2017-07-03 paper/content/parallelHaskells.tex
added olegl changes

\end{flushright}\section{Background}
\label{sec:background}
\subsection{Short introduction to parallel Haskells}
\label{sec:parEvalNIntro}
There are already several ways to write parallel programs in Haskell. As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel or |parEvalN :: [a -> b] -> [a] -> [b]|, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of |parEvalN|.}
	\label{fig:parEvalN}
\end{figure}
Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement |parEvalN| with several different parallel Haskells.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|.

For |parEvalN| this means that we can just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator. We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form. To ensure that programs that use |parEvalN| have the correct evaluation order, we annotate the computation with |pseq :: a -> b -> b| which forces the compiler to not reorder multiple |parEvalN| computations. This is particularly necessary in circular communication topologies like in the |torus| or |ring| (Chap.~\ref{sec:topology-skeletons}), where a wrong execution order would result in deadlock scenarios when executed without |pseq|.

The resulting code and a graphical representation can be found in Fig.~\ref{fig:parEvalNMulticore} and Fig.~\ref{fig:parEvalNMulticoreImg}, respectively.

\begin{figure}[h]
\begin{code}
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = let bs = zipWith ($) fs as 
                 in (bs `using` parList rdeepseq) `pseq` bs
\end{code}
\caption{Multicore version of |parEvalN|.}
\label{fig:parEvalNMulticore}
\end{figure}

\begin{figure}[h]
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
	\caption{Dataflow of the Multicore Haskell |parEvalN| version.}
	\label{fig:parEvalNMulticoreImg}
\end{figure} %$ %% formatting

\subsubsection{ParMonad}
The |Par| monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a monad designed for composition of parallel programs.

The |Par| Monad version of our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| monad with |get|. This results in |Par [b]|. We execute this process with |runPar| to finally get the final |[b]|.

Again, the resulting code and a graphical representation can be found in Fig.~\ref{fig:parEvalNParMonad} and Fig.~\ref{fig:parEvalNParMonadImg}, respectively.

@@

\begin{figure}[h]
\begin{code}
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = runPar $ 
	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
\end{code}
\caption{|Par| Monad version of |parEvalN|.}
\label{fig:parEvalNParMonad}
\end{figure}
\begin{figure}[h]
	\includegraphics[scale=0.5]{images/parEvalNParMonad}
	\caption{Dataflow of the |Par| Monad |parEvalN| version.}
	\label{fig:parEvalNParMonadImg}
\end{figure}

@@

\begin{code}
parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
parEvalN = spawnF 
\end{code}
\begin{figure}[h]
	\includegraphics[scale=0.5]{images/parEvalNEden}
	\caption{Dataflow of the Eden |parEvalN| version.}
	\label{fig:parEvalNEden}
\end{figure}
A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.

\paragraph{Eden TraceViewer.}
\label{sec:edentv}
To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}. In the next sections we will present some \emph{traces}, the post-mortem process diagrams of Eden processes and their activity.

In a trace, the $x$ axis shows the time, the $y$ axis enumerates the machines and processes. A~trace shows a running process in green, a blocked process is red. If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow. The typical reason for then is GC. An inactive machine where no processes are started yet, or all are already terminated, is shows as a blue bar. A~comminication from one process to another is represented with a black arrow. A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes. An example trace can be found in Fig.~\ref{fig:withoutFutures}.

>>>>

\end{flushright}\section{Background}
\label{sec:background}
\subsection{Short introduction to parallel Haskells}
\label{sec:parEvalNIntro}
There are already several ways to write parallel programs in Haskell. As we base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel or |parEvalN :: [a -> b] -> [a] -> [b]|, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of |parEvalN|.}
	\label{fig:parEvalN}
\end{figure}
%Before we go into detail on how we can use this idea of parallelism
%for parallel Arrows, as a short introduction to parallelism in
%Haskell we will now implement |parEvalN| with several different
%parallel Haskells.
As a demonstration, we implement here the non-Arrows |parEvalN| in multiple parallel Haskells.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1998a} is a way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|. Let:

\begin{code}
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = let bs = zipWith ($) fs as 
                 in (bs `using` parList rdeepseq) `pseq` bs
\end{code} %$

In the above definition of |parEvalN| we just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator. We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form. To ensure that programs that use |parEvalN| have the correct evaluation order, we annotate the computation with |pseq :: a -> b -> b| which forces the compiler to not reorder multiple |parEvalN| computations. This is particularly necessary in circular communication topologies like in the |torus| or |ring| (Chap.~\ref{sec:topology-skeletons}), where a wrong execution order would result in deadlock scenarios when executed without |pseq|.
\olcomment{CAREFUL! is |pseq| REALLY about multiple instances or \emph{merely} order of execution?}
%
Fig.~\ref{fig:parEvalNMulticoreImg} shows a visual representation of this code.

% \begin{figure}[h]
% \begin{code}
% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
% parEvalN fs as = let bs = zipWith ($) fs as 
%                  in (bs `using` parList rdeepseq) `pseq` bs
% \end{code}
% \caption{Multicore version of |parEvalN|.}
% \label{fig:parEvalNMulticore}
% \end{figure}

\begin{figure}[t]
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
	\caption{Dataflow of the Multicore Haskell |parEvalN| version.}
	\label{fig:parEvalNMulticoreImg}
\end{figure} %$ %% formatting

\subsubsection{ParMonad}
The |Par| monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a monad designed for composition of parallel programs. Let:

\begin{code}
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = runPar $ 
	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
\end{code}

The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| monad with |get|. This results in |Par [b]|. We execute this process with |runPar| to finally get the final |[b]|.
A graphical representation can be found Fig.~\ref{fig:parEvalNParMonadImg}.

@@

% \begin{figure}[h]
% \begin{code}
% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
% parEvalN fs as = runPar $ 
% 	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
% \end{code}
% \caption{|Par| Monad version of |parEvalN|.}
% \label{fig:parEvalNParMonad}
% \end{figure}
\begin{figure}[t]
	\includegraphics[scale=0.5]{images/parEvalNParMonad}
	\caption{Dataflow of the |Par| Monad |parEvalN| version.}
	\label{fig:parEvalNParMonadImg}
\end{figure}

@@

\begin{code}
parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
parEvalN = spawnF 
\end{code}
% \begin{figure}[h]
% 	\includegraphics[scale=0.5]{images/parEvalNEden}
% 	\caption{Dataflow of the Eden |parEvalN| version.}
% 	\label{fig:parEvalNEden}
% \end{figure}
% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.
\olcomment{removed the image as it shows nothing new}

\paragraph{Eden TraceViewer.}
\label{sec:edentv}
To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}. In the next sections we will present some \emph{trace visualizations}, the post-mortem process diagrams of Eden processes and their activity.

The trace visualizations are color-coded. In such a visualization (Fig.~\ref{fig:withoutFutures}), the $x$ axis shows the time, the $y$ axis enumerates the machines and processes. The visualization shows a running process in green, a blocked process is red. If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow. The typical reason for thus is GC. An inactive machine, where no processes are started yet, or all are already terminated, shows as a blue bar. A~comminication from one process to another is represented with a black arrow. A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.

### 2017-07-03 paper/content/parrows.tex
added olegl changes

\section{Parallel Arrows}
\label{sec:parallel-arrows}
We have seen what Arrows are and how they can be used as a general interface to computation. In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textit{parallel computation} as well. We start by introducing the interface and explaining the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\subsection{The ArrowParallel typeclass}
As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions |a -> b| in parallel, |parEvalN| (Chap.~\ref{sec:parEvalNIntro}).
Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
This combinator is similar to our utility function |listApp| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
% \begin{figure}[h]
\begin{code}
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator as a function}
% \label{fig:parEvalNArrowFn}
% \end{figure}
With this definition of |parEvalN|, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new typeclass |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
% \begin{figure}[h]
\begin{code}
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClass1}
% \end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak head normal form vs. normal form). For this reason we also introduce an additional |conf| parameter to the function. We also do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|. So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
% \begin{figure}[h]
\olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
\begin{code}
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClassFinal}
% \end{figure}
Note that we don't require the |conf| parameter in every implementation. If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|, as we will see in the implementation of the Multicore Haskell and the |Par| Monad backends.

@@

\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
\begin{figure}[t]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}% $ %% formatting
\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this. We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|: 
% \begin{figure}[h]
\begin{code}
data Conf a = Conf (Strategy a)
\end{code}
% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}
We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances. To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[h]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}
\caption{Configurable ArrowParallel instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{|Par| Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting}
The |Par| monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| monad's |spawnP :: NFData a => a -> Par (IVar a)| function. The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| monad.

We therefore apply each function to its corresponding input value with || and then fork the computation away with |arr spawnP| inside a |zipWithArr| (Fig.~\ref{fig:zipWithArr}) call. This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|. In order to wait for the computation to finish, we map over the |IVar|s inside the |Par| monad with |arr (>>= mapM get)|. The result of this operation is a |Par [b]| from which we can finally remove the monad again by running |arr runPar| to get our output of |[b]|.
\begin{figure}[h]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = 
			(arr $ \as -> (fs, as)) >>>
			zipWithArr (app >>> arr spawnP) >>>
			arr sequenceA >>>
			arr (>>= mapM get) >>>
			arr runPar
\end{code} %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend.}
\label{fig:ArrowParallelParMonad}
\end{figure}

\subsubsection{Eden}
For both the Multicore Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| typeclasses. With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows. While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
% \begin{figure}[h]
\begin{code}
class (Arrow arr) => ArrowUnwrap arr where
	arr a b -> (a -> b)
\end{code}
% \caption{possible ArrowUnwrap typeclass}
% \label{fig:ArrowUnwrap}
% \end{figure}
We don't do it here for aesthetic resons, though. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
% \begin{figure}[h]
\begin{code}
instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
parEvalN _ fs as = spawnF fs as
\end{code}
% \caption{ArrowParallel instance for functions in the Eden backend}
% \label{fig:ArrowParallelEdenFns}
% \end{figure}
and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
% \begin{figure}[h]
\begin{code}
instance (Monad m, Trans a, Trans b, Trans (m b)) =>
	ArrowParallel (Kleisli m) a b conf where
parEvalN conf fs =
	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
	(Kleisli $ sequence)
\end{code}
% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
% \label{fig:ArrowParallelEdenKleisli}
% \end{figure}

>>>>

\section{Parallel Arrows}
\label{sec:parallel-arrows}
Arrows are a general interface to computation. Here we introduce special Arrows as general interface to \textit{parallel computations}. First, we present the interface and explain the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\subsection{The ArrowParallel typeclass}
A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Sec.~\ref{sec:parEvalNIntro}).
Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
This combinator is similar to our utility function |listApp| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
% \begin{figure}[h]
\begin{code}
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator as a function}
% \label{fig:parEvalNArrowFn}
% \end{figure}
With this definition of |parEvalN|, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new typeclass |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
% \begin{figure}[h]
\begin{code}
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClass1}
% \end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak head normal form vs. normal form). For this reason we also introduce an additional |conf| parameter to the function. We also do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|. % So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
% % \begin{figure}[h]
% \olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
\begin{code}
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{code}
% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClassFinal}
% \end{figure}
We do not require the |conf| parameter in every implementation. If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|.
% , as we will see in the implementation of the Multicore Haskell and the |Par| Monad backends.

@@

\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
\begin{figure}[t]
\begin{code}
data Conf a = Conf (Strategy a)

instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}% $ %% formatting
\caption{Fully evaluating |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this. We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
% \begin{figure}[h]
% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}
% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances. To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[t]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}
\caption{Configurable |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{|Par| Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"?}
The |Par| monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| monad's |spawnP :: NFData a => a -> Par (IVar a)| function. The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| monad.

We therefore apply each function to its corresponding input value with and then fork the computation away with |arr spawnP| inside a |zipWithArr| (Fig.~\ref{fig:zipWithArr}) call. This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|. In order to wait for the computation to finish, we map over the |IVar|s inside the |Par| monad with |arr (>>= mapM get)|. The result of this operation is a |Par [b]| from which we can finally remove the monad again by running |arr runPar| to get our output of |[b]|.
\begin{figure}[h]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = 
			(arr $ \as -> (fs, as)) >>>
			zipWithArr (app >>> arr spawnP) >>>
			arr sequenceA >>>
			arr (>>= mapM get) >>>
			arr runPar
\end{code} %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend.}
\label{fig:ArrowParallelParMonad}
\end{figure}

\subsubsection{Eden}
For both the Multicore Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| typeclasses. With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows. While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
% \begin{figure}[h]
\begin{code}
class (Arrow arr) => ArrowUnwrap arr where
	arr a b -> (a -> b)
\end{code}
% \caption{possible ArrowUnwrap typeclass}
% \label{fig:ArrowUnwrap}
% \end{figure}
% We don't do it here
However, we avoid doing so for aesthetic resons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
% \begin{figure}[h]
\begin{code}
instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
parEvalN _ fs as = spawnF fs as
\end{code}
% \caption{ArrowParallel instance for functions in the Eden backend}
% \label{fig:ArrowParallelEdenFns}
% \end{figure}
and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
% \begin{figure}[h]
\begin{code}
instance (Monad m, Trans a, Trans b, Trans (m b)) =>
	ArrowParallel (Kleisli m) a b conf where
parEvalN conf fs =
	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
	(Kleisli $ sequence)
\end{code}
% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
% \label{fig:ArrowParallelEdenKleisli}
% \end{figure}
